---
title: "ECO 530 - Poster Draft 2"
author: "Fall 2024 - Ibrahim Quansah"
output:
  html_document: default
  word_document: default
  pdf_document: default
---



```{r, warning=FALSE, message=FALSE}

#Loading Libraries

library(tidyverse)
library(kableExtra)
library(vtable)
library(modelsummary)
library(dplyr)
library(readr)
library(modelsummary)
library(ggplot2)
library(knitr)
library(tinytex)
library(tseries)
library(estimatr)
library(car)
library(lmtest)
```



**Research Question**

Is there any relationship between U.S. economic growth (GDP) and trade volumes (Net Trade)?

Purpose: To identify whether a correlation exists between the two variables, meaning that changes in trade volumes might correlate with changes in economic growth (GDP).


**Introduction**

This study explores the relationship between economic growth, measured by Gross Domestic Product (GDP), and trade volumes, specifically net trade, in the United States. Net trade refers to the difference between a country’s exports and imports, and it plays a crucial role in shaping the overall economic performance of a nation. In the context of the U.S., a key player in global trade, understanding how changes in trade volumes influence economic growth is critical for evaluating economic policies and trade agreements.

I will utilize data on U.S. GDP and Net Trade from the World Bank DataBank, covering two decades. The GDP represents the total economic output of the country, while net trade captures the balance of goods and services traded internationally.



**Data**

```{r, warning=FALSE, message=FALSE}
#Reading Net Trade File
Net_Trade <- read_csv("~/ECO 530/Data/Net Trade.csv")
head(Net_Trade)

```

```{r, warning=FALSE, message=FALSE}
#Reading GDP File
GDP <- read_csv("~/ECO 530/Data/GDP.csv")
head(GDP)
```

```{r, warning=FALSE, message=FALSE}
#Combining GDP and Net Trade in one data
Data <- merge(Net_Trade, GDP, by = 'DATE', all = TRUE)

Data$DATE <- as.Date(Data$DATE, format = "%m/%d/%Y")

# Sort chronologically
Data <- Data %>% arrange(DATE)

#Formatting Table with Kable
Data %>%
  kable(caption = "U.S. GDP and Net Trade (1993-2023)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 
```

```{r, warning=FALSE, message=FALSE}

# Summary statistics for GDP and Net Trade

Summary_Table <- Data %>%
  summarise(
    GDP_Mean = mean(GDP),
    GDP_Median = median(GDP),
    GDP_SD = sd(GDP),
    Net_Trade_Mean = mean(`Net Trade`),
    Net_Trade_Median = median(`Net Trade`),
    Net_Trade_SD = sd(`Net Trade`)
  )

# Creating a Table
Table <- Summary_Table %>%
  pivot_longer(cols = everything(), names_to = "Metric", values_to = "Value")

# Displaying the table with kableExtra
Table %>%
  kable("html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

Table
```


# Creating a line plot for GDP and Net Trade over time
```{r, warning=FALSE, message=FALSE}
# Melt the data to long format for ggplot2
Data$DATE <- as.Date(Data$DATE, format = "%m/%d/%Y")
Data$Years <- format(Data$DATE, "%Y")
Data_long <- Data %>%
  pivot_longer(cols = c(GDP, `Net Trade`), names_to = "Variable", values_to = "Value") %>%
  select(Years, Variable, Value)

# Create the line plot
ggplot(Data_long, aes(x = Years, y = Value, color = Variable, group = Variable)) +
  geom_line() +
  geom_point() +
  labs(title = "GDP and Net Trade Over Time", x = "Year", y = "Value (Billion USD)") +
  scale_y_continuous(labels = scales::comma, breaks = seq(0, 30000000, 5000000))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
```



```{r, warning=FALSE, message=FALSE}
#Creating Scatter Plot of Net Trade and GDP
ggplot(Data, aes(x = GDP, y = `Net Trade`)) +
  geom_point(color = "blue", size = 3) +  # Points in blue
  labs(title = "Scatter Plot of Net Trade vs. GDP",
       x = "GDP (Billion USD)",
       y = "`Net Trade` (Billion USD)") +
  theme_minimal() +  # Use a minimal theme
  scale_x_continuous(labels = scales::comma)+
  geom_smooth(method = "lm", color = "red", se = FALSE)


# Calculate correlation between GDP and Net Trade
correlation <- cor(Data$GDP, Data$`Net Trade`)

# Display the correlation value
correlation
```

```{r, warning=FALSE, message=FALSE}
#Reading GDPDEF File
GDPDEF <- read_csv("~/ECO 530/Data/GDPDEF.csv")

#Reading Interest Rate File
Interest_Rate <- read_csv("~/ECO 530/Data/FEDFUNDS.csv")


#Reading Government Expenditure File
Gov_Expen <- read_csv("~/ECO 530/Data/Gov Expen.csv")
colnames(Gov_Expen)[colnames(Gov_Expen) == "W068RCQ027SBEA"] <- "Government Expenditure"


#Reading Unemployment Rate File
UNRate <- read_csv("~/ECO 530/Data/UNRate.csv")


#Reading Gross Private Domestic Investment (GPDI) File
GPDI <- read_csv("~/ECO 530/Data/GPDI.csv")

```


```{r, warning=FALSE, message=FALSE}
Data2 <- data.frame(
  DATE = Data$DATE,  # Assuming all lists have the same years
  Net_Trade = Net_Trade$`Net Trade`, 
  GDP = GDP$GDP,
  GDPDEF = GDPDEF$GDPDEF,
  Interest_Rate = Interest_Rate$'FEDFUNDS',
  Gov_Expen = Gov_Expen$'Government Expenditure',
  UNRate = UNRate$"UNRATE",
  GPDI = GPDI$GPDI
)
# Sort chronologically
Data2 <- Data2 %>% arrange(DATE)
```


```{r, warning=FALSE, message=FALSE}
#Creating the lagged variables 
Data2 <- Data2 %>% 
  mutate( Lagged_GDP = dplyr::lag(GDP, 1), 
          Lagged_Net_Trade = dplyr::lag(Net_Trade, 1) )
```

```{r, warning=FALSE, message=FALSE}
# Specify the file path and name 
file_path <- "Data2.csv" # Create the CSV file 
write.csv(Data2, file_path, row.names = FALSE)
```

```{r, warning=FALSE, message=FALSE}
Poster_Data <- read_csv("~/ECO 530/Data/Data2.csv")
# Convert the 'DATE' column to a proper date format
Poster_Data$DATE <- as.Date(Poster_Data$DATE, format = "%m/%d/%Y") 

# Create a 'Years' column by extracting the year from the 'DATE'
Poster_Data$Years <- as.numeric(format(Poster_Data$DATE, "%Y"))
```

```{r, warning=FALSE, message=FALSE}
#Performing the ADF test on GDP and Nettrade 
adf.test(Poster_Data$GDP) 
adf.test(Poster_Data$Net_Trade)
```

```{r, warning=FALSE, message=FALSE}
# List of control variables to check 
control_vars <- c("GDP", "GDPDEF", "Interest_Rate", "Gov_Expen", "UNRate", "GPDI") 
# Function to apply ADF test and print results 
check_stationarity <- function(var_name, data) { test_result <- adf.test(data[[var_name]], alternative = "stationary") 
print(paste("ADF Test for", var_name, ": p-value =", test_result$p.value)) } 
# Check stationarity for each control variable 
lapply(control_vars, check_stationarity, data = Poster_Data)
```

```{r, warning=FALSE, message=FALSE}
reg1 <- lm_robust (Lagged_GDP ~ Lagged_Net_Trade, Poster_Data)


modelsummary(list("Model 1" =reg1) , gof_map = c("nobs"), stars = TRUE,
             fmt = 2, statistic= 'std.error', 
             notes ="Robust SEs in parentheses",
             title = "Regression Analysis of Net Trade on GDP",
             output="kableExtra") %>% kable_classic_2()
```


```{r, warning=FALSE, message=FALSE}
reg2 <- lm_robust (Lagged_GDP ~ Lagged_Net_Trade + Years, Poster_Data)

modelsummary(list("Model 2" = reg2), 
              gof_map = c("nobs"), 
              stars = TRUE,
              statistic = 'std.error', 
              notes = "Robust SEs in parentheses",
              title = "Regression Analysis of Net Trade on GDP",
              output = "kableExtra") %>% 
  kable_classic_2()

```

```{r, warning=FALSE, message=FALSE}
reg3 <- lm_robust(Lagged_GDP ~ Lagged_Net_Trade + Years + GDPDEF + Interest_Rate + Gov_Expen + UNRate + GPDI, data = Poster_Data)


modelsummary(list("Model 3"=reg3) , gof_map = c("nobs"), stars = TRUE,
              statistic= 'std.error', 
             notes ="Robust SEs in parentheses", 
             output="kableExtra") %>% kable_classic()

```

```{r, warning=FALSE, message=FALSE}
calculate_vif <- function(reg3) {
  formula <- formula(reg3)
  variables <- all.vars(formula)[-1]  # Exclude the dependent variable
  vif_values <- sapply(variables, function(var) {
    other_vars <- variables[variables != var]
    sub_model <- lm(as.formula(paste(var, "~", paste(other_vars, collapse = "+"))), data = Poster_Data)
    1 / (1 - summary(sub_model)$r.squared)
  })
  return(vif_values)
}

# Apply custom VIF calculation to lm_robust
vif_values <- calculate_vif(reg3)
print(vif_values)
```

```{r, warning=FALSE, message=FALSE}
dwtest(reg3)
```

```{r, warning=FALSE, message=FALSE}
cor(Poster_Data[, c("Lagged_GDP", "Lagged_Net_Trade", "Years", "GDPDEF", "Interest_Rate", "Gov_Expen", "UNRate", "GPDI")])

```

```{r, warning=FALSE, message=FALSE}

# Ensure Poster_Data is sorted by 'Years' to maintain order
Poster_Data <- Poster_Data %>%
  arrange(Years)

# Create differenced variables, filling the first value with NA
Poster_Data <- Poster_Data %>%
  mutate(
    GDP_diff = c(NA, diff(GDP)),
    NetTrade_diff = c(NA, diff(Net_Trade)),
    GDPDEF_diff = c(NA, diff(GDPDEF)),
    Interest_Rate_diff = c(NA, diff(Interest_Rate)),
    Gov_Expen_diff = c(NA, diff(Gov_Expen)),
    UNRate_diff = c(NA, diff(UNRate)),
    GPDI_diff = c(NA, diff(GPDI)),
  )

# Remove the first row with NA values due to differencing
Poster_Data <- na.omit(Poster_Data)

# Check the structure of the filtered dataset to confirm changes
str(Poster_Data)



```


```{r, warning=FALSE, message=FALSE}
# List of differenced variables to check
differenced_vars <- c("GDP_diff", "GDPDEF_diff", "Interest_Rate_diff", "Gov_Expen_diff", "UNRate_diff", "GPDI_diff")

# Check stationarity for each differenced variable
lapply(differenced_vars, check_stationarity, data = Poster_Data)

```

```{r, warning=FALSE, message=FALSE}
# Run the regression model with the differenced variables
reg4 <- lm_robust(
  GDP_diff ~ NetTrade_diff + Years + GDPDEF_diff + Interest_Rate_diff + Gov_Expen_diff + UNRate_diff + GPDI_diff,
  data = Poster_Data
)

# Summarize the regression results
modelsummary(list("Model 4"=reg4) , gof_map = c("nobs"), stars = TRUE,
              statistic= 'std.error', 
             notes ="Robust SEs in parentheses", 
             output="kableExtra") %>% kable_classic()
```

```{r, warning=FALSE, message=FALSE}
calculate_vif <- function(reg4) {
  formula <- formula(reg4)
  variables <- all.vars(formula)[-1]  # Exclude the dependent variable
  vif_values <- sapply(variables, function(var) {
    other_vars <- variables[variables != var]
    sub_model <- lm(as.formula(paste(var, "~", paste(other_vars, collapse = "+"))), data = Poster_Data)
    1 / (1 - summary(sub_model)$r.squared)
  })
  return(vif_values)
}

# Apply custom VIF calculation to lm_robust
vif_values2 <- calculate_vif(reg4)
print(vif_values2)

```
